---
biblio-style: apalike
highlight_bw: yes
output:
  bookdown::pdf_document2:
    toc: false
    includes:
      in_header: header.tex
    dev: "cairo_pdf"
    latex_engine: xelatex
    fig_caption: yes
geometry: margin=0.45in
link-citations: yes
---

**Supplemental File of**

\begingroup\Large
**clusterProfiler 4.0: A universal enrichment tool for functional and comparative study**
\endgroup

**Tianzhi Wu§, Erqiang Hu§, Meijun Chen, Pingfan Guo, Zehan Dai, Tingze Feng, Shuangbin Xu, Lang Zhou, Wenli Tang, Li Zhan, Xiaocong Fu, Shanshan Liu, Xiaochen Bo\* and Guangchuang Yu\* **

^\*^correspondence: Guangchuang Yu \<gcyu1@smu.edu.cn\> and Xiaochen Bo \<boxc@bmi.ac.cn\>


\renewcommand{\figurename}{Fig.}
\newcommand{\beginsupplement}{%
\setcounter{table}{0}
\renewcommand{\thetable}{S\arabic{table}}%
\setcounter{figure}{0}
\renewcommand{\thefigure}{S\arabic{figure}}%
}

\beginsupplement

```{r, echo=FALSE, message=FALSE, results='hide'}
require(kableExtra)
options(knitr.table.fromat = "latex")
knitr::opts_chunk$set(fig.pos= "!ht")
#knitr::opts_chunk$set(out.width="90%")
knitr::opts_chunk$set(fig.align="center")
usepackage_latex("float")
usepackage_latex("makecell")
usepackage_latex("booktabs")
```

```{r, echo=FALSE, message=FALSE}
library(DOSE)
library(clusterProfiler)
library(enrichplot)
```

# R packages that depends on clusterProfiler


The `clusterProfiler` library is one of the fundamental packages and it had been incorporated in more than twenty R packages (in CRAN or Bioconductor) to perform functional enrichment analysis for different topics, especially for cancer research.

```{r}
db <- utils::available.packages(repo=BiocManager::repositories())
pkgs <- tools::package_dependencies('clusterProfiler', db=db, 
                            which = c("Depends", "Imports"), 
                            reverse=TRUE)[[1]]
pkgs                                                                      
```

```{r}
library(kableExtra)
d <- data.frame(Package = pkgs,
Description = sapply(pkgs, function(p) gsub("\n", " ", packageDescription(p)$Title)
))
caption <- "R packages that rely on clusterProfiler to perform functional analysis."
knitr::kable(d, caption = caption)
```


# Examples of using clusterProfiler


**NOTE:** source codes and datasets to produce this file can be obtained online^[<https://github.com/YuLab-SMU/supplemental-clusterProfiler-v4>].

# References
